Safe Transferable Regions

### 2017-03-22 paper/ecoop_rebuttal.txt

We thank the reviewers for their extensive review of our work, and
thoughtful suggestions for improvement.

1. Why are closures treated specially? Why don't classes suffice?
It is true that C# treats lambdas as objects of class `Func`. A flip
side of such punning is that C# classes cannot have type-polymorphic
lambdas as class fields (since all fields have monomorphic types).
Adopting C#'s approach in the region type system would lead to a
similar drawback: we can no longer support region-polymorphic lambdas
as class fields, but this feature is used frequently in practice. An
example is presented in Fig.3 (and explained in 3rd paragraph on p.8).

2. We are summarily in agreement with reviewer 24A's comparison with
HM. Separation of type and region parameters is an example of the
language aspect designed with inference in mind. "Action at distance
errors" do occur in our case, and we conjecture that the recent
advances[oopsla15,pldi15] could help us migitage the problem.

3. Reviewer 24B's IntBox example is (rightfully) rejected by the type
system. The type system enforces two invariants on every well-typed
object: (a). its type signature explicitly captures the regions to
which its fields refer to, and (b). all such regions must outlive the
allocation region of the object. The discussion on constrained region
polymorphism (p.8) covers this informally, but formal enforcement is
done by the class well-formedness rule, which was relegated to the
appendix (Fig. 13). In the context of reviewer’s example, if
`b.boxref` is a reference to an object in region `R2`, and if `b`
itself (an `IntBoxBox`) is allocated in `R1`, then the type signature
is `IntBoxBox<R1,R2 | R2≥R1>`. The example (implicitly) tries to
instantiate `R1` with `x` (rather, it’s region idenfitier), and `R2`
with `y`, but note that there does not exist any outlives relationship
between R1 and R2 (since both are transferable regions). Thus, this
instantiation is disallowed, effectively preempting the possibility of
an unsafe reference.

4. Reviewer 24B's lambda example is (incorrectly) accepted by the type
system. The reason for this is that the closure type rule of Fig. 6 is
missing the premise `Δ ≥ r` (Δ outlives r). We had this premise in the
previous draft of the paper, but got rid of it during the days leading
to the submission, perhaps reasoning (incorrectly) that it is a
tautology. We regret this decision. Restoring this premise would
prevent the example from type checking since `x` doesn't outlive `y`.

5. It is true that the FB^+ language used in the evaluation is
improverished in comparison with Java/C#, but is is enough to capture
the aspects of Naiad operators relevant to the region type system.
There are a total of 6 Naiad streaming query operators, out of which
we analyzed 4 (Select, SelectMany, Join and WindowJoin). The other two
are aggregation operators, which do not contain any interesting use
of regions. Instances of the safety violation described in the paper
were found in all 4, and the violation can manifest in an actual
execution in in each case. We accept the reviewer's suggestion to
include these details in the paper.

### 2017-03-22 paper/ecoop_rebuttal.txt

We thank the reviewers for their extensive review of our work, and
thoughtful suggestions for improvement.

1. Why are closures treated specially? Why don't classes suffice?
It is true that C# treats lambdas as objects of class `Func`. A flip
side of such punning is that C# classes cannot have type-polymorphic
lambdas as class fields (since all fields have monomorphic types).
Adopting C#'s approach in the region type system would lead to a
similar drawback: we can no longer support region-polymorphic lambdas
as class fields, but this feature is used frequently in practice. An
example is presented in Fig.3 (and explained in 3rd paragraph on p.8).

2. We agree with reviewer 24A's comparison with HM. Separation of type
and region parameters is an example of the language aspect designed with
inference in mind. "Action at distance errors" do occur in our case,
and we conjecture that the recent advances[oopsla15,pldi15] could help
us migitage the problem.

3. Reviewer 24B's IntBox example is (rightfully) rejected by the type
system. The type system enforces two invariants on every well-typed
object: (a) Its type signature explicitly captures the regions to
which its fields refer to, and (b) All such regions must outlive the
allocation region of the object. The discussion on constrained region
polymorphism (p.8) covers this informally, but formal enforcement is
done by the class well-formedness rule, which was (unfortunately) relegated to the
appendix (Fig. 13). In the context of reviewer’s example, if
`b.boxref` is a reference to an object in region `R2`, and if `b`
itself (an `IntBoxBox`) is allocated in `R1`, then the type signature
is `IntBoxBox<R1,R2 | R2≥R1>`. The example (implicitly) tries to
instantiate `R1` with `x` (rather, it’s region idenfitier), and `R2`
with `y`, but note that there does not exist any outlives relationship
between R1 and R2 (since both are transferable regions). Thus, this
instantiation is disallowed, effectively preempting the possibility of
an unsafe reference.

4. (Lambda allocation regions) Reviewer 24B's intuition and points in
this regard are correct. The closure type rule of Fig. 6 is
unfortunately missing a premise, namely `Δ ≥ r` (every region in Δ outlives r).
Adding this premise will ensure that the reviewer's example is correctly handled
(since `x` doesn't outlive `y`). And it is also correct that with this
premise, the allocation of lambda does not have to be restricted to the
allocation region.

We had an equivalent premise in the previous draft of the paper, but got rid
of it during the days leading to the submission, with a faulty simplification
considering only allocation regions. We apologize for the confusion caused by this mistake.

### 2017-03-22 paper/ecoop_rebuttal.txt

We thank the reviewers for their extensive review of our work, and
thoughtful suggestions for improvement.

1. Why are closures treated specially? Why don't classes suffice?
It is true that C# treats lambdas as objects of class `Func`. A flip
side of such punning is that C# classes cannot have type-polymorphic
lambdas as class fields (since all fields have monomorphic types).
Adopting C#'s approach in the region type system would lead to a
similar drawback: we can no longer support region-polymorphic lambdas
as class fields, but this feature is used frequently in practice. An
example is presented in Fig.3 (and explained in 3rd paragraph on p.8).

2. We agree with reviewer 24A's comparison with HM. Separation of type
and region parameters is an example of the language aspect designed with
inference in mind. "Action at distance errors" do occur in our case,
and we conjecture that the recent advances[oopsla14,pldi15] could help
us migitage the problem.

@@

3. Reviewer 24B's IntBox example is (rightfully) rejected by the type
system. The type system enforces two invariants on every well-typed
object: (a) Its type signature explicitly captures the regions to
which its fields refer to, and (b) All such regions must outlive the
allocation region of the object. The discussion on constrained region
polymorphism (p.8) covers this informally, but formal enforcement is
done by the class well-formedness rule, which was (unfortunately) relegated to the
appendix (Fig. 13). In the context of reviewer’s example, if
`b.boxref` is a reference to an object in region `R2`, and if `b`
itself (an `IntBoxBox`) is allocated in `R1`, then the type signature
is `IntBoxBox<R1,R2 | R2≥R1>`. The example (implicitly) tries to
instantiate `R1` with `x` (rather, it’s region idenfitier), and `R2`
with `y`, but note that there does not exist any outlives relationship
between R1 and R2 (since both are transferable regions). Thus, this
instantiation is disallowed, effectively preempting the possibility of
an unsafe reference.

4. (Lambda allocation regions) Reviewer 24B's intuition and points in
this regard are correct. The closure type rule of Fig. 6 is
unfortunately missing a premise, namely `Δ ≥ r` (every region in Δ outlives r).
With this added premise, the reviewer's example is correctly handled
(since `x` doesn't outlive `y`). And it is also correct that with this
premise, the allocation of lambda does not have to be restricted to the
allocation region.

We had an equivalent premise in the previous draft of the paper, but
got rid of it during the days leading to the submission, with a faulty
simplification considering only allocation regions. We apologize for
the confusion caused by this mistake. Thanks to the reviewer for
catching this.

@@

5. It is true that the FB^+ language used in the evaluation is
improverished in comparison with Java/C#, but is is enough to capture
the aspects of Naiad operators relevant to the region type system.
There are a total of 6 Naiad streaming query operators, out of which
we analyzed 4 (Select, SelectMany, Join and WindowJoin). The other two
are aggregation operators, which do not contain any interesting use
of regions. Instances of the safety violation described in the paper
were found in all 4, and the violation can manifest in an actual
execution in in each case. We accept the reviewer's suggestion to
include these details in the paper.

6. In the definition of C-decomposable, why only one of the regions is
required to be in Δ? 
The asymmetry in this definition reflects properties of the
constraint-generation, namely that the context Φ is modified only by
adding conjuncts of the form `Δ ≥ π` where `π` is not a
region-parameter (see the rule LetRegion), as discussed in the proof
of Lemma 3 in Appendix C.

7. In the explanation of the type safety result, the statement "the
theorem also captures the key property of operational semantics that
no live region is ever freed or transferred" is informal for "if
`consistent(Δ,Σ)` and `(e,Σ)-->(e',Σ')` then `consistent(Δ,Σ')`"
(Lemma 8 of the appendix). Thus, operational semantics in fact never
frees or transfers a live region. We however understand that the
informal statement can be misconstrued as stronger guarantees provided
by the type statement, and accept the reviewer's suggestion to include
a clarification.

### References ###

[oopsla14]: Z Pavlinovic, Finding minimum type error sources, OOPSLA'14
[pldi15]: D Zhang et al, Diagnosing type errors with class, PLDI'15

### 2017-03-22 paper/ecoop_rebuttal.txt

We thank the reviewers for their extensive review of our work, and
thoughtful suggestions for improvement.

1. Why are closures treated specially? Why don't classes suffice?
It is true that C# treats lambdas as objects of class `Func`. A flip
side of such punning is that C# classes cannot have type-polymorphic
lambdas as class fields (since all fields have monomorphic types).
Adopting C#'s approach in the region type system would lead to a
similar drawback: we can no longer support region-polymorphic lambdas
as class fields, but this feature is used frequently in practice. An
example is presented in Fig.3 (and explained in 3rd paragraph on p.8).

2. We agree with reviewer 24A's comparison with HM. Separation of type
and region parameters is an example of the language aspect designed with
inference in mind. "Action at distance errors" do occur in our case,
and we conjecture that the recent advances[oopsla14,pldi15] could help
us mitigate the problem.

3. Reviewer 24B's IntBox example is (rightfully) rejected by the type
system. The type system enforces two invariants on every well-typed
object: (a) Its type signature explicitly captures the regions to
which its fields refer to, and (b) All such regions must outlive the
allocation region of the object. The discussion on constrained region
polymorphism (p.8) covers this informally, but formal enforcement is
done by the class well-formedness rule, which was (unfortunately) relegated to the
appendix (Fig. 13). In the context of reviewer’s example, if
`b.boxref` is a reference to an object in region `R2`, and if `b`
itself (an `IntBoxBox`) is allocated in `R1`, then the type signature
is `IntBoxBox<R1,R2 | R2≥R1>`. The example (implicitly) tries to
instantiate `R1` with `x` (rather, it’s region identifier), and `R2`
with `y`, but note that there does not exist any outlives relationship
between R1 and R2 (since both are transferable regions). Thus, this
instantiation is disallowed, effectively preempting the possibility of
an unsafe reference.

@@

4. (Lambda allocation regions) Reviewer 24B's intuition and points in
this regard are correct. The closure type rule of Fig. 6 is
unfortunately missing a premise, namely `Δ ≥ r` (every region in Δ outlives r).
With this added premise, the reviewer's example is correctly handled
(since `x` doesn't outlive `y`). And it is also correct that with this
premise, the allocation of lambda does not have to be restricted to the
allocation region.

We had an equivalent premise in the previous draft of the paper, but
got rid of it during the days leading to the submission, with a faulty
simplification considering only allocation regions. We apologize for
the confusion caused by this mistake. Thanks to the reviewer for
catching this.

5. It is true that the FB^+ language used in the evaluation is
impoverished in comparison with Java/C#, but is is enough to capture
the aspects of Naiad operators relevant to the region type system.
There are a total of 6 Naiad streaming query operators, out of which
we analyzed 4 (Select, SelectMany, Join and WindowJoin). The other two
are aggregation operators, which do not contain any interesting use
of regions. Instances of the safety violation described in the paper
were found in all 4, and the violation can manifest in an actual
execution in in each case. We accept the reviewer's suggestion to
include these details in the paper.

@@

6. In the definition of C-decomposable, why only one of the regions is
required to be in Δ? 
The asymmetry in this definition reflects properties of the
constraint-generation, namely that the context Φ is modified only by
adding conjuncts of the form `Δ ≥ π` where `π` is not a
region-parameter (see the rule LetRegion), as discussed in the proof
of Lemma 3 in Appendix C.

7. In the explanation of the type safety result, the statement "the
theorem also captures the key property of operational semantics that
no live region is ever freed or transferred" is informal for "if
`consistent(Δ,Σ)` and `(e,Σ)-->(e',Σ')` then `consistent(Δ,Σ')`"
(Lemma 8 of the appendix). Thus, operational semantics in fact never
frees or transfers a live region. We however understand that the
informal statement can be misconstrued as stronger guarantees provided
by the type system. We accept the reviewer's suggestion to include
a clarification.

### 2017-03-22 paper/ecoop_rebuttal.txt

We thank the reviewers for their extensive review of our work, and
thoughtful suggestions for improvement.

1. Why are closures treated specially? Why don't classes suffice?
It is true that C# treats lambdas as objects of class `Func`. A flip
side of such punning is that C# classes cannot have type-polymorphic
lambdas as class fields (since all fields have monomorphic types,
i.e., there cannot be any quantifiers).  Adopting C#'s approach in the
region type system would lead to a similar drawback: we can no longer
support region-polymorphic lambdas as class fields, but this feature
is used frequently in practice. An example is presented in Fig.3 (and
explained in 3rd paragraph on p.8).

### 2017-03-22 paper/ecoop_rebuttal.txt

5. It is true that the FB^+ language used in the evaluation is
impoverished in comparison with Java/C#, but is is enough to capture
the aspects of Naiad operators relevant to the region type system.
There are a total of 6 Naiad streaming query operators, out of which
we analyzed 4 (Select, SelectMany, Join and WindowJoin). The other two
are aggregation operators, which do not contain any interesting use
of regions. Instances of the safety violation described in the paper
were found in all 4, and the violation can manifest in an actual
execution in in each case. We accept the reviewer's suggestion to
include these details in the paper.

6. In the definition of C-decomposable, why only one of the regions is
required to be in Δ? 
The asymmetry in this definition reflects properties of the
constraint-generation, namely that the context Φ is modified only by
adding conjuncts of the form `Δ ≥ π` where `π` is not a
region-parameter (see the rule LetRegion), as discussed in the proof
of Lemma 3 in Appendix C.

7. In the explanation of the type safety result, the statement "the
theorem also captures the key property of operational semantics that
no live region is ever freed or transferred" is informal for "if
`consistent(Δ,Σ)` and `(e,Σ)-->(e',Σ')` then `consistent(Δ,Σ')`"
(Lemma 8 of the appendix). Thus, operational semantics in fact never
frees or transfers a live region. We however understand that the
informal statement can be misconstrued as stronger guarantees provided
by the type system. We accept the reviewer's suggestion to include
a clarification.

8. Why do we need constraint based lifetime parameters for classes?
What are the distinctive examples which this design enables which a
simpler ownership system cannot provide? 
Transferable regions are self-contained, containing no objects with
pointers to objects in other regions. However, stack regions are more
flexible, and are permitted to contain objects with pointers to
objects in other regions. Stack regions are very useful as "working
storage" (paragraph 3 on p.3). For example, consider a dataflow
computation that indexes the intermediary results obtained from an
upstream actor in a transferable region (`R1`), before updating the
results, and transferring them to a downstream actor. The index,
perhaps a B-tree, is an auxiliary structure that need not be
transferred, hence can be stored in a stack region (`R2`). Outlives
constraints are needed to admit safe references from `R2` into `R1`.

### 2017-03-22 paper/ecoop_rebuttal.txt

We thank the reviewers for their extensive review of our work, and
thoughtful suggestions for improvement.

1. **Why are closures treated specially? Why don't classes suffice?**
It is true that C# treats lambdas as objects of class `Func`. A flip
side of such punning is that C# classes cannot have type-polymorphic
lambdas as class fields (since all fields have monomorphic types,
i.e., there cannot be any quantifiers).  Adopting C#'s approach in the
region type system would lead to a similar drawback: we can no longer
support region-polymorphic lambdas as class fields, but this
functionality is required in practice to support user-defined
functions (UDFs) in LINQ queries. An example is presented in Fig.3
(and explained in 3rd paragraph on p.8).

2. We agree with reviewer 24A's **comparison with HM**. Separation of type
and region parameters is an example of the language aspect designed
with inference in mind. "Action at distance errors" do occur in our
case, and we conjecture that the recent advances
[[oopsla14][oopsla14],[pldi15][pldi15]] could help us mitigate the
problem.

3. Reviewer 24B's **`IntBox` example** is (rightfully) rejected by the
type system. The type system enforces two invariants on every
well-typed object: (a) Its type signature explicitly captures the
regions to which its fields refer to, and (b) All such regions must
outlive the allocation region of the object. The discussion on
constrained region polymorphism (p.8) covers this informally, but
formal enforcement is done by the class well-formedness rule, which
was (unfortunately) relegated to the appendix (Fig. 13). In the
context of reviewer’s example, if `b.boxref` is a reference to an
object in region `R2`, and if `b` itself (an `IntBoxBox`) is allocated
in `R1`, then the type signature is `IntBoxBox<R1,R2 | R2≥R1>`. The
example (implicitly) tries to instantiate `R1` with (the region
identifier of) `x`, and `R2` with `y`, but note that there does not
exist any outlives relationship between R1 and R2 (since both are
transferable regions). Thus, this instantiation is disallowed,
effectively preempting the possibility of an unsafe reference.

4. **Lambda allocation regions**: Reviewer 24B's intuition and points
in this regard are correct. The closure type rule of Fig. 6 is
unfortunately missing a premise, namely `Δ ≥ r` (every region in Δ
outlives r). The premise requires a closure to be well-typed under
only the regions that outlive its allocation region. With this added
condition, the reviewer's example is correctly handled (since `x`
doesn't outlive `y`). And it is also correct that with this premise,
the allocation of lambda does not have to be restricted to the
allocation region.
We had an equivalent premise in the previous draft of the paper, but
got rid of it during the days leading to the submission, with a faulty
simplification considering only allocation regions. We apologize for
the confusion caused by this mistake. We are grateful to the reviewer
for pointing it out.

5. It is true that the FB^+ **language used in the evaluation** is
impoverished in comparison with Java/C#, but is is enough to capture
the aspects of Naiad operators relevant to the region type system.
There are a total of 6 Naiad streaming query operators, out of which
we analyzed 4 (Select, Join, GroupBy, and Aggregate). The other two
are not interesting from the region type system perspective. Instances
of the safety violation described in the paper were found in all 4,
and the violation can manifest in an actual execution in in each case.
We accept reviewer 24B's suggestion to include these details in the
paper. To answer reviewer 24D's question of whether the need to clone
objects negates the benefits of regions, we believe the answer is no.
As explained in paragraph 2 on p.1 (citing [7]), the savings obtained
by using regions are primarily because of (a). avoiding multiple older
generation collections over long-lived objects, and (b). preventing
cascading effect of long GC pauses on the latency at downstream actors.
Both these factors are unaffected by cloning, which happens only once
per objects (as against multiple GC walks over a large data structure
that aggregates objects).

6. **In the definition of C-decomposable, why only one of the regions is
required to be in Δ?**
The asymmetry in this definition reflects properties of the
constraint-generation, namely that the context `Φ` is modified only by
adding conjuncts of the form `Δ ≥ π` where `π` is not a
region-parameter (see the rule LetRegion), as discussed in the proof
of Lemma 3 in Appendix C.

7. In the **explanation of the type safety result**, the statement
"the theorem also captures the key property of operational semantics
that no live region is ever freed or transferred" is informal for "if
`consistent(Δ,Σ)` and `(e,Σ)-->(e',Σ')` then `consistent(Δ,Σ')`"
(included in the type safety statement, and proved as Lemma 8 in the
appendix). Thus, operational semantics in fact never frees or
transfers a live region. We however understand that the informal
statement can be misconstrued as stronger guarantees provided by the
type system. We accept the reviewer's suggestion to include a
clarification.

8. **Why do we need constraint based lifetime parameters for classes?
What are the distinctive examples which this design enables which a
simpler ownership system cannot provide?**  Transferable regions are
self-contained, containing no objects with pointers to objects in
other regions. However, stack regions are more flexible, and are
permitted to contain objects with pointers to objects in other
regions. Stack regions are very useful as "working storage" (paragraph
3 on p.3). For example, consider a dataflow computation that indexes
the intermediary results obtained from an upstream actor in a
transferable region (`R1`), before updating the results, and
transferring them to a downstream actor. The index, perhaps a B-tree,
is an auxiliary structure that need not be transferred, hence can be
stored in a stack region (`R2`). Outlives constraints are needed to
admit safe references from `R2` into `R1`.

[oopsla14]: http://dl.acm.org/citation.cfm?id=2660230
[pldi15]: http://dl.acm.org/citation.cfm?id=2738009

### 2018-03-20 paper/ecoop18_rebuttal.txt

We thank the reviewers for their reviews and suggestions for improvement.

Review #50A
------------

1. The other two Naiad benchmarks that were left out of evaluation
were so left out because they are very similar to the existing
benchmarks insofar as region type system is concerned. In particular,
the Naiad vertex `RegionWindowJoinVertex` is similar to
`RegionJoinVertex`, and `RegionSelectManyVertex` is similar to
`RegionSelectVertex`. They were nonetheless included in the evaluation
of [7] because they allow the runtime overhead of regions to be
measured at scale (allowing [7] to make a case for regions from the
performance standpoint, as opposed to the safety standpoint). We shall
include this note in the final version of the paper.

Review #50B
-------------

1. Dealing with run-time errors: Run-time exceptions are raised either
when the program attempts to transfer/free an open region, or open a
transferred/freed region (The operational semantics (Figs. 15 and 16)
precisely identify these two cases). The former is often a mere
programming oversight, which can be rectified by moving the
`transfer`/`free` out of the lexical scope of `open`. The latter could
however denote a genuine bug in the program, where a dynamic region is
assumed to be always available at some program point, when in fact
it could have been freed/transferred along a program path leading to
that point. The solution in this case is to fix the program logic by
acknowledging that the region may no longer be available at that
program point.

Review #50C
-----------

1. How the type system differentiates between `free` and `transfer`:
The type system doesn't. Dynamic regions in Broom are primarily
intended to transfer data among actors (hence the name "Transferable
Regions"). It is therefore expected that a dynamic region is
transferred at least once during its lifetime (before it is freed),
thereby making it unsafe to allow outgoing references from the region
to other regions.  While nothing prevents a transferable region to be
used merely as dynamic memory, that is not an intended usecase for
which Broom is optimized.

2. Serializing: Broom's region runtime [7] serializes data if the
receiver of a transferable region is remote.

3. Our model does admit aliases (through the `let` expression). The
operational semantics defines a a *store* (Σ) that maps locations to
their typestates, and implicitly models a heap at the granularity of a
region (each location $l$ denotes a region, and every object whose
first region parameter is $l$ (e.g., `new B<l,..>)(...)`) is allocated
in the region at $l$). Mutable object fields constitute an ostensible
omission from our model (also the FGJ model), but this isn't a major
shortcoming considering that constructor application (during the new
object creation) includes assignments to object fields. We would
include a detailed note to this extent in the final version.

4. As far as the syntax is concerned, `transfer` and `free` are like
any other methods, hence their application is modeled by the generic
method application expression ($e.m<...>(...)$). Since the operational
semantics models a single actor (and not a system of actors), the
behavior of `free` is same as that of `transfer`: both result in the
region becoming unavailable to the actor calling `transfer`/`free`.
The rule for `transfer` captures this semantics.

5. The run-time checks are precisely captured by the premises of
the opsem rules that deal with exceptions (namely, OpenTransferred and
TransferOpened). The checks basically check whether the action being
performed (e.g., `transfer`) is allowed given the current typestate of
the region (e.g., "live"), raising an exception otherwise.

6. `inRgn.free()` in Fig. 3 is needed to prevent a memory leak.
Reference going out of scope doesn't automatically free a region,
because the region may have been aliased. Naiad code does contain
`free` statements.

### 2018-03-20 paper/implementation.tex
Ecoop18-related changes and the rebuttal.

We performed two kinds of experiments to evaluate our region type system and type
inference.
First, we implemented some 
microbenchmarks ($\le$100 LOC) consisting of standard classes such
as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
our inference engine to infer their region types.
These classes are region-oblivious. Hence, as long as they are well-typed
as per the core type system, \namec must be able to automatically construct its
region-type-annotated definition without fail. \namec was
able to infer the expected region types for all these classes under 10ms.
Fig~\ref{fig:rev} shows the region-type-annotated
definition computed for the list reverse method. Observe that \namec
was able to infer that the list and its data (of type \C{T}) can be
allocated in different regions, as long as the later outlives the
former. This allows, for instance, a \C{preOrder} method to traverse a
tree in a transferable region, and return a list of its nodes, where
the list itself is allocated in the stack region.

@@

Next, we translated some of the Naiad streaming query operator
benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
$\absof{\FB^+}$, and used \namec to verify their safety.
The number of LOC performing operations on \C{Region} objects relative
to the total LOC is 8\% or under in the Naiad benchmarks. 
During the process, we found multiple instances of potential memory safety
violations in the $\absof{FB^+}$ translation of benchmarks, which we
verified to be present in the original C\# implementation as well. The
cause of all safety violations is the creation of a reference from the
outgoing message (a transferable region) to the payload of the
incoming message. For example, the implementation of \C{SelectVertex}
contains the following:
\begin{codejava}
  if (this.selector(inMsg.payload[i])) {
    outMsg.set(outputOffset, inMsg.payload[i]);
    ...
  } 
\end{codejava}
The \C{outMsg} is later transferred to a downstream actor, where the
reference to \C{inMsg}'s payload becomes unsafe\footnote{This unsafe
reference could have gone unnoticed during experiments
in~\cite{Broom:HotOS} because their experimental setup included only
one actor.}. We eliminated such unsafe references by creating a clone
of \C{inMsg.payload[i]} in \C{outMsg}, and our compiler was
subsequently able to certify the safety of all references.

>>>>

We performed two kinds of experiments to evaluate our region type system and type
inference.
First, we implemented some 
microbenchmarks ($\le$100 LOC) consisting of standard classes such
as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
our inference engine to infer their region types.
These classes are region-oblivious. Hence, as long as they are well-typed
as per the core type system, \namec must be able to automatically construct its
region-type-annotated definition without fail. \namec was
able to infer the expected region types for all these classes under 10ms.
Fig~\ref{fig:rev} shows the region-type-annotated
definition computed for the list reverse method. Observe that \namec
was able to infer that the list and its data (of type \C{T}) can be
allocated in different regions, as long as the latter outlives the
former. This allows, for instance, a \C{preOrder} method to traverse a
tree in a transferable region, and return a list of its nodes, where
the list itself is allocated in the stack region.

@@

Next, we translated 4 out of the 6 Naiad streaming query operator
benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
$\absof{\FB^+}$, and used \namec to verify their safety (The remaining
two Naiad vertices are not interesting from the type system
perspective).  The number of LOC performing operations on \C{Region}
objects relative to the total LOC is 8\% or under in the Naiad
benchmarks.  During the process, we found multiple instances of
potential memory safety violations in the $\absof{FB^+}$ translation
of all the 4 Naiad vertices, which we verified to be present in the
original C\# implementation as well. The cause of all safety
violations is the creation of a reference from the outgoing message (a
transferable region) to the payload of the incoming message. For
example, the implementation of \C{SelectVertex} contains the
following:
\begin{codejava}
  if (this.selector(inMsg.payload[i])) {
    outMsg.set(outputOffset, inMsg.payload[i]);
    ...
  } 
\end{codejava}
The \C{outMsg} is later transferred to a downstream actor, where the
reference to \C{inMsg}'s payload becomes unsafe\footnote{This unsafe
reference could have gone unnoticed during experiments
in~\cite{Broom:HotOS} because their experimental setup included only
one actor.}. We eliminated such unsafe references by creating a clone
of \C{inMsg.payload[i]} in \C{outMsg}, and our compiler was
subsequently able to certify the safety of all references.

### 2018-03-20 paper/introduction.tex
Sec.3 edits complete.
Ecoop18-related changes and the rebuttal.

\mypara{Language.}
The cornerstone of this approach is an \C{open} lexical block for transferable regions,
that ``opens'' a transferable region and guarantees that the region
won't be transferred/freed while it is open. By nesting a
Tofte-Talpin style \C{letregion} lexical block, that delimits
the lifetime of a stack region, inside an \C{open} lexical block for a
transferable region, we can guarantee that the transferable region
will remain live as long as the stack region is live. We say that the
former \emph{outlives} the later., and any references from the stack
region to the transferable region are therefore safe.
This is particular useful, since such stack regions serve as temporary
working storage while working with the (opened) transferable regions.

@@

\item \broom{} is equipped with a region type system that statically
  guarantees safety of all memory accesses in a well-typed program,
  provided that certain typestate invariants on regions hold.  The
  later invariants are enforced via simple runtime checks.

>>>>

\mypara{Language.}
The cornerstone of this approach is an \C{open} lexical block for transferable regions,
that ``opens'' a transferable region and guarantees that the region
won't be transferred/freed while it is open. By nesting a
Tofte-Talpin style \C{letregion} lexical block, that delimits
the lifetime of a stack region, inside an \C{open} lexical block for a
transferable region, we can guarantee that the transferable region
will remain live as long as the stack region is live. We say that the
former \emph{outlives} the latter, and any references from the stack
region to the transferable region are therefore safe. This is
particular useful, since such stack regions serve as temporary working
storage while working with the (opened) transferable regions.

@@

\item \broom{} is equipped with a region type system that statically
  guarantees safety of all memory accesses in a well-typed program,
  provided that certain typestate invariants on regions hold.  The
  latter invariants are enforced via simple runtime checks.

### 2018-03-20 paper/related.tex
Ecoop18-related changes and the rebuttal.

Our region type system can also be thought of as a specialized
ownership type system~\cite{OwnershipSurvey}, where each region is the
owner of all objects allocated in the region.
%
An ownership type system for safe region-based memory management in
real-time Java has been proposed by~\cite{MIT03}.  
% Like us, they too assume a source language with programmer-managed
% memory regions, and focus on proving safety of programs written in
% that language.  
Their language permits only lexically-scoped (stack) regions.
% (in the context of shared-memory concurrency).
In contrast, we permit regions with dynamically determined lifetimes.
% (in the context of message-passing concurrency).
% We borrow outlives relation from their formal
% development, and our type system bears some similarities to theirs.
Our language also admits generics and higher-order functions.
We also establish type safety and transfer safety results that
formalize the guarantees provided by our system. While~\cite{MIT03}'s
language is explicitly typed, our language comes equipped with full
type inference.
However, several inference algorithms have been proposed in the context
of other ownership type systems.
Our type inference algorithm is also novel compared to these existing
ownership inference algorithms, which are based on, e.g.,
pointer analysis~\cite{HuangEtAl:ECOOP12} or boolean satisfiability~\cite{DietlEtAl:ECOOP11}.
(See Section 5.2 of~\cite{OwnershipSurvey} for a more comprehensive discussion
of ownership inference algorithms.)
Some distinguishing characteristics of our algorithm is that it is customized
to our problem, does not use any pointer analysis algorithm (which
can be a source of imprecision) or SAT solvers (which can be a source of inefficiency),
and comes with relative completeness guarantees.

~\cite{HMN01} proposes a flow-sensitive approach for first-order
programs to generalize~\cite{tofte97}'s approach to dynamic
regions.~\cite{CR04} describes a flow-insensitive and
context-sensitive analysis that transforms Java programs to use
(dynamic) regions.  However, neither~\cite{HMN01} nor~\cite{CR04}
supports dynamic regions as first-class objects; they cannot be stored
in data structures or passed to methods. Furthermore,
while~\cite{HMN01} requires reference counting to ensure memory
safety,~\cite{CR04} comes with no formal safety guarantees.
%
\cite{gpu14} uses regions to safely transfer data between the CPU and GPU
in the context of Scheme. However, their setting only includes lexically-scoped
regions for which Tofte and Talpin-style analysis~\cite{tofte97}
suffices. In contrast, we provide first-class support for
transferable regions with dynamic lifetimes.
% We require this generality in order to support streaming query operators, such as the
% one shown in Fig.~\ref{fig:motivating-eg}.

>>>>

Our region type system can also be thought of as a specialized
ownership type system~\cite{OwnershipSurvey}, where each region is the
owner of all objects allocated in the region.
%
An ownership type system for safe region-based memory management in
real-time Java has been proposed by Boyapati \emph{et al.}~\cite{MIT03}.  
% Like us, they too assume a source language with programmer-managed
% memory regions, and focus on proving safety of programs written in
% that language.  
Their language permits only lexically-scoped (stack) regions.
% (in the context of shared-memory concurrency).
In contrast, we permit regions with dynamically determined lifetimes.
% (in the context of message-passing concurrency).
% We borrow outlives relation from their formal
% development, and our type system bears some similarities to theirs.
Our language also admits generics and higher-order functions.  We also
establish type safety and transfer safety results that formalize the
guarantees provided by our system. While Boyapati \emph{et al.}'s
language is explicitly typed, our language comes equipped with full
type inference.  However, several inference algorithms have been
proposed in the context of other ownership type systems.  Our type
inference algorithm is also novel compared to these existing ownership
inference algorithms, which are based on, e.g., pointer
analysis~\cite{HuangEtAl:ECOOP12} or boolean
satisfiability~\cite{DietlEtAl:ECOOP11}.  (See Section 5.2 of Clark
\emph{et al.}'s survey of ownership type
systems~\cite{OwnershipSurvey} for a more comprehensive discussion of
ownership inference algorithms.) Some distinguishing characteristics
of our algorithm is that it is customized to our problem, does not use
any pointer analysis algorithm (which can be a source of imprecision)
or SAT solvers (which can be a source of inefficiency), and comes with
relative completeness guarantees.

Henglein \emph{et al.}~\cite{HMN01} propose a flow-sensitive approach
for first-order programs to generalize Tofte and Talpin's approach to
dynamic regions. Cherem and Rugina~\cite{CR04} describe a
flow-insensitive and context-sensitive analysis that transforms Java
programs to use (dynamic) regions.  However, neither of themr supports
dynamic regions as first-class objects; they cannot be stored in data
structures or passed to methods. Furthermore, while Henglein \emph{et
al.}~\cite{HMN01} require reference counting to ensure memory safety,
Cherem and Rugina's analysis~\cite{CR04} comes with no formal safety
guarantees.
%
Holk \emph{et al.}~\cite{gpu14} use regions to safely transfer data
between the CPU and GPU in the context of Scheme. However, their
setting only includes lexically-scoped regions for which Tofte and
Talpin-style analysis suffices. In contrast, we provide first-class
support for transferable regions with dynamic lifetimes.
% We require this generality in order to support streaming query operators, such as the
% one shown in Fig.~\ref{fig:motivating-eg}.

### 2018-03-20 paper/type-inference.tex
Ecoop18-related changes and the rebuttal.

Theorem~\ref{thm:constraint-generation-sc}
and~\ref{thm:constraint-solver-sc} show that the second and third
steps of the type-inference algorithm are sound. It is easy to verify
the soundness of the first step (the elaboration phase): For any
program $p \in \absof{\FB}$, the elaboration phase produces a $q$ such
that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.
%
The soundness of the type inference algorithm follows.

>>>>

Theorems~\ref{thm:constraint-generation-sc}
and~\ref{thm:constraint-solver-sc} show that the second and third
steps of the type-inference algorithm are sound. It is easy to verify
the soundness of the first step (the elaboration phase): For any
program $p \in \absof{\FB}$, the elaboration phase produces a $q$ such
that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.
%
The soundness of the type inference algorithm follows.
